### 2019-03-24 paper/sex_selection_spacing-ver2.tex
Minor edits after final read-through

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference.
In India, the sex of the last-born child significantly impacts the duration to next birth,
with the expected birth interval about three percent longer if the last-born was a son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam the more boys a family has the longer is the expected 
duration to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have a substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health---especially 
for very short intervals of 18 months or less---although identifying the causal mechanisms 
have proven more difficult \citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There are three pathways through which birth intervals can affect health.
First, there is maternal depletion, where the mother does not have sufficient time to 
recover after her last pregnancy before she becomes pregnant again.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiency.
}
Second, disease transmission either from mother to child or between children.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is of particular interest when son preference drives spacing 
decisions.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to affect the older child negatively.

In India, there is evidence that both the prior and the subsequent child have worse health 
outcomes and a higher likelihood of dying the shorter the interval between them, although 
the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008,Ghosh2018}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explains a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again sooner, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, El Salvadore, and Pakistan that
short intervals between births have adverse health effects
\citep{Cleland1984,Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

@@

Second, have the shapes of the survival curves changed?
Very short birth intervals are associated with worse health outcomes and increased 
mortality risk \citep{Conde-Agudelo2012}.
Hence, understanding how the distribution of birth intervals changes help us understand to 
what extent we can expect improvements in health outcomes.
We should, for example, expect different impacts on the distribution of birth intervals
depending on whether parents change their spacing behavior based on an increasing 
concern for child health or from a strong son preference leading to continued pregnancies
until they conceive a son.

Finally, how are the changes in birth intervals with increasing access to sex selection 
distributed geographically?
Given the size and cultural differences within India, it is unlikely that the effect of
increasing access to sex selection will have had the same effect throughout 
\citep{retherford03b}.
I, therefore, examine how birth intervals have changed across four broad regions.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differentially.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing to the birth of a girl or a boy.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that the 
effect of household characteristics such as the sex composition of previous births have 
the same effect throughout the entire spell as is assumed by the proportional hazard
model.
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth 
and how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.

The explanatory variables in $\mathbf{Z}$ and the interactions between them 
are the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard and $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.%
\footnote{
I discuss the choice of which variables to use for non-proportionality in more detail
in the ``Explanatory Variables'' section.
}

@@

I exclude visitors to the household, as well as
women who have been married more than once, divorced, or who are 
not living with their husband,
women with inconsistent age at marriage,
and those with missing information on education.
Women interviewed in NFHS-3 or NFHS-4 who were never married or where Gauna 
had not yet been performed were also dropped.
The same goes for women who had at least one multiple births,
reported having a birth before age 12, had a birth before marriage, or
duration between births of less than nine months.

@@

% [How to read tables]
Figures \ref{fig:bootstrap_low}, \ref{fig:bootstrap_med}, and  \ref{fig:bootstrap_high} 
show predicted average birth intervals, sex ratios, and probabilities of having
a birth by decade, spell, and sex composition for the three education levels separated 
by the area of residence.
For each subsample, I first estimate the model described in Section 
\ref{sec:strategy} and then use the estimated coefficients to predict 
average spell length, sex ratio, and the probability of having a birth in that spell.

@@

For legibility, none of the graphs show standard errors.
Instead, the standard errors, together with the graphed values, are shown in Appendix 
Tables 
\ref{tab:bootstrap_duration_avg_sex_ratio_low_all}, 
\ref{tab:bootstrap_duration_avg_sex_ratio_med_all}, and 
\ref{tab:bootstrap_duration_avg_sex_ratio_high_all}.
The standard errors are based on bootstrapping for all three measures,
where the model is repeatedly estimated using resampling with replacement.

@@

Where son preference in birth intervals shows most clearly is for the fourth spell.
Despite the increase in spacing even for those with three girls, the increases in 
spacing have been substantially stronger for families with at least one son.
As for fertility, the more sons a family has, the more significant are the differences.

@@

Figure \ref{fig:bootstrap_med} shows that women with one to seven years of education 
follow a fertility pattern broadly similar to those with no education.
As expected, fertility is lower than for women with no education, but still a
strong son preference in the parity progression behavior.
Except for the second spell, the likelihood of having a birth by the end of the spell is
substantially lower if there has been at least one male birth than if there are only girls.
Despite this, the probability of a birth in the third and fourth spells has also been 
falling for those with only girls.
Having a second birth is still close to universal, although in urban areas it is down to
just over 85 percent for those with a son as their first child.

@@

The rapid decline in fertility for the most educated women is even more clear for the 
third and fourth spells when the woman has given birth to at least one son.
At the beginning of the sample, urban women with at least one boy had a close to 75
percent likelihood of having a birth, but that has fallen to only around 25 percent.
However, son preference still shows clearly in the parity progression probability for 
women without a son, with the decline for the third and fourth spell substantially
slower for both urban and rural women without a son than for those with at least one son.
The falling fertility means that relatively few women make it to the fourth spell,
which, therefore, has more noise than the other spells.

@@

The tremendous impact that sex selection can have on birth spacing is illustrated 
particularly vividly by the changes over time for the third spell.
In the 1972-1984 period, the predicted average birth intervals for the three possible sex 
compositions were less than two months apart.
For the later periods, however, the highest predicted average spacing was for women
with only daughters with most of the differences to the other sex compositions 
statistically significant.
In urban areas, the predicted average birth interval for a woman with only daughters 
increased by almost a full year from 26.8 to 38.3 months.
What is more, this increase is likely an underestimate.
The sex ratios for the 1972-84 period are substantially higher than the natural sex ratio 
despite sex selection not being widely available until around 1985.
If this early, elevated sex ratio is because of recall error, it means that the actual birth 
intervals would have been substantially shorter than the ones predicted here and the
increase over time larger.

The reversal pattern also shows for the fourth spell for urban women, although not for
rural women.
An important caveat is that because of the falling fertility relatively few women make
it to the fourth spell, and even if they do, we do not observe a birth by the time of the
survey.
Of the 11,886 women across urban and rural areas observed for the fourth spell for the 
2005-2016 period more than 70 percent did not have a birth at the time of the survey.
Even though the censoring is less prevalent for the earlier periods, this is countered by
those samples being substantially smaller.
Despite this, the pattern for the fourth spell is similar to that of the third spell.

@@

The results above show consistent increases in average spacing across the board.
Average spacing is a convenient way to understand the overall changes in behavior but may 
hide important differences in the distribution of spacing.
Specifically, whether these increases in average spacing come from fewer births after 
very short intervals, because of even longer spacing from births with already long 
spacing, or a general increase in spacing have important implications for what effects the
increase in average spacing are likely to have and how to interpret the motivation for
the changes.
On one hand, if there still are many births that occur after very short intervals, the 
increase in average spacing is unlikely to have a substantial impact on health outcomes
since the literature on birth spacing and child health suggests that an interpregnancy 
interval of fewer than 18 months is associated with a higher risk of adverse health 
outcomes and mortality \citep{Conde-Agudelo2012}.
On the other hand, we may see a reduction in early observed births and more long spells if 
families with fewer sons are both more likely to have a very early pregnancy and to 
continue through pregnancies and sex-selective abortions until they conceive a son.%
\footnote{
It is unclear what the optimal response in the timing of pregnancies to the availability 
of sex selection should be.
The expected birth intervals become longer with the use of sex selection, which may be 
costly for parents leading them to try to conceive even earlier than previously.
Conversely, if parents know that short birth intervals are detrimental to the next child's
health, and with sex selection, the next child is likely a boy, it might make sense to 
wait longer before conceiving.
}

@@

Figure \ref{fig:pps_low} shows the distribution of birth within a spell for women with no 
education.
For the second spell, what is most interesting is that there is almost no difference 
between those with a first-born boy and those with a first-born girl.
For the 1972--1984 period the two conditional survival curves only begin to diverge 
at around 18 months (corresponding to 27 months after the first birth took place), and
at that point close to 50 percent of the women have already had their second child.
With a nine-month pregnancy, this means that almost 50 percent of the children born are
born within the interpregnancy interval associated with a higher risk of adverse health
outcomes and mortality, and, somewhat surprisingly, that there do not appear to be any 
particular disadvantage for the child born after a girl than after a boy.%
\footnote{
Note that this does not imply that the first-born girl is not in a more adverse position
than a first-born boy when the second child arrives, just that this is not directly
related to a short birth interval.
}
The difference in spacing between a first-born girl and a first-born boy first-born 
remains almost the same across the four decades.
Furthermore, the proportion with very short spacing remains almost the same.

For the third spell, the son preference driven short spacing is evident throughout the
distribution and becomes more pronounced over the four decades.
Families with two girls are especially likely to have a birth early with 25 percent
having a child within 12 months of the beginning of the spell (21 months after the prior
second birth).
That number is only slightly smaller four decades later. 
The fourth spell shows changes in the distribution that appear to be more pronounced
than for the third spell, although these results are based on fewer observed births.
For families with three girls, 75 percent have had their fourth child within 36 months 
after the beginning of the spell, but this occurs more than a year later for families
with two or more boys.

@@

For the third spell, the 1972--1984 period show a close to standard pattern throughout, 
although, as mentioned above, the sex ratios for women without girls was substantially 
higher than the natural sex ratio either because of recall error or early use of 
sex selection.
The correlation between the increased use of sex selection and the reversal of the 
standard spacing pattern show up clearly in the 2005--2016 period for the third spell with 
spacing to the third child consistently longer with two girls than with either two boys 
or one boy and one girl throughout the entire spell.
This pattern is most consistent with early and continued use of sex selection, rather 
than a general reduction in the likelihood of very short birth intervals.
It also means, that previous girls can expect more time without a younger sibling,
which may reduce the sibling competition for resources.

@@

Figure \ref{fig:bootstrap_high_r2} shows the same outcomes for the ``North'' region. 
Despite the idea that the ``North'' has high fertility and lower 
likelihood of sex selection the differences to the ``West'' region is surprisingly small,
at least in the urban areas.
Fertility is, indeed, higher than in the ``West'' region, but the declines in the 
probability of parity progression have been relatively rapid and mirror the national one.
The use of sex selection is also close to the ``West,'' and there is the same reversal of
spacing pattern as there.
The rural areas of the ``North'' are a slightly different story.
Fertility is substantially higher, and the use of sex selection much lower than in the
rural areas of the ``West.'' 
That said, there still is clear evidence that the sex ratio is substantially above normal
at close to 60 percent for the latest decade.

@@

The final region is the ``South,'' shown in Figure \ref{fig:bootstrap_high_r4}.
This region is historically thought to have the lowest degree of son preference of the
regions of India.
As for the ``East,'' I do not show the fourth spell results because of the small sample 
size and the associated noisy estimates.
Judging from the second spell progression rate and the associated spacing pattern, it
does, indeed, appear that there is little son preference in this region.
The likelihood of having a second birth is relatively high at around 80 to 90 percent
and is close to independent of the sex of the first birth in both urban and rural areas.
The predicted average spacing also show no sign of son preference, and the predicted
sex ratios are close to the natural rate.
The average birth spacing started high relative to the other regions, and with little 
change in fertility and no significant use of sex selection, it has increased only 
slightly over time.

With the high likelihood of a second birth and lower decline in the likelihood of a third
birth, the ``South'' has higher predicted fertility than the ``East.''
What is more, for urban areas the average birth interval has increased substantially
for those with only girls relative to those with a boy and a girl, corresponding to
a predicted sex ratio close to 60 percent with only girls. 
The signs of the use of sex selection also show up for rural women with no sons.
Furthermore, over the four decades, the likelihood of having a third birth has decreased
much less than for either a boy and a girl or two boys.%
\footnote{
The relatively high parity progression for women without son does raise the question of
whether ``daughter avoidance'' is behind the fertility behavior in this region as
suggested by \citet{DiamondSmith2008}.
}
The differences in parity progression and the sex ratios suggest that despite originally
having less son preference, the ``South'' now appears to have longer average spacing 
in the absence of sons and the higher sex ratio associated with the use of sex selection.

@@

At the other extreme, urban women with eight or more years of education have experienced 
an almost complete reversal of the traditional spacing patterns as a result of falling
fertility and the associated extensive use of sex selection.
Rather than having the next birth sooner as they mostly did before sex selection became 
available, women with either no or one son now have substantially longer spacing than if 
they have two or more sons.
As shown by the increasingly significant deviations in the sex ratio from the natural sex 
ratio, these changes are likely due to increased use of sex selection rather than a sudden
reversal in son preference.
Only the second spell has not seen a reversal despite a substantially increased use of
sex selection, mainly because the average spacing with a first-born boy has increased 
independently by more than nine months.

>>>>

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference.
In India, the sex of the last-born child has in the past significantly affected the 
duration to next birth, with the expected birth interval about three percent longer if the 
last-born was a son than if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam, the more boys a family has the longer is the expected 
duration to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have a substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health---especially 
for very short intervals of 18 months or less---although identifying the causal mechanisms 
have proven more difficult \citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There are three pathways through which birth intervals can affect health.
First, there is maternal depletion, where the mother does not have sufficient time to 
recover after her last pregnancy before she becomes pregnant again.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiencies.
}
Second, disease transmission either from mother to child or between children.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is of particular interest when son preference drives spacing 
decisions.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to affect the older child negatively.

In India, there is evidence that both the prior and the subsequent child have worse health 
outcomes and a higher likelihood of dying the shorter the interval between them, although 
the effects are not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008,Ghosh2018}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explains a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again sooner, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, El Salvadore, and Pakistan that
short intervals between births have adverse health effects
\citep{Cleland1984,Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

@@

Second, have the shapes of the survival curves changed?
Very short birth intervals are, as mentioned, associated with worse health outcomes and 
increased mortality risk \citep{Conde-Agudelo2012}.
Hence, understanding how the distribution of birth intervals changes can help us understand to 
what extent we can expect improvements in health outcomes.
We should, for example, expect different impacts on the distribution of birth intervals
depending on whether parents change their spacing behavior based on an increasing 
concern for child health or from a strong son preference leading to continued pregnancies
until they conceive a son.

Finally, how are the changes in birth intervals with increasing access to sex selection 
distributed geographically?
Given the size and cultural differences within India, it is unlikely that the increasing 
access to sex selection will have had the same effects throughout \citep{retherford03b}.
I, therefore, examine how birth intervals have changed across four regions.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing to the birth of a girl or a boy.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth 
and how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from nine months after one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.%
\footnote{
I discuss the choice of which variables to use for non-proportionality in more detail
in the ``Explanatory Variables'' section.
}

@@

I exclude visitors to the household, as well as
women who have been married more than once, divorced, or who are 
not living with their husband,
women with inconsistent age at marriage,
and those with missing information on education.
Women interviewed in NFHS-3 or NFHS-4 who were never married or where Gauna 
had not yet been performed were also dropped.
The same goes for women who had at least one multiple births,
reported having a birth before age 12, had a birth before marriage, or an interval between 
births of less than nine months.

@@

% [How to read tables]
Figures \ref{fig:bootstrap_low}, \ref{fig:bootstrap_med}, and  \ref{fig:bootstrap_high} 
show predicted average birth intervals, sex ratios, and probabilities of having
a birth by decade, spell, and sex composition for the three education levels separated 
by the area of residence.%
\footnote{
For legibility, none of the graphs show standard errors.
Instead, the standard errors, together with the graphed values, are shown in Appendix 
Tables 
\ref{tab:avg_sex_ratio_low}, 
\ref{tab:avg_sex_ratio_med}, and 
\ref{tab:avg_sex_ratio_high}.
The standard errors are based on bootstrapping for all three measures,
where the model is repeatedly estimated using resampling with replacement.
}
For each subsample, I first estimate the model described in Section 
\ref{sec:strategy} and then use the estimated coefficients to predict 
average birth spacing, sex ratio, and the probability of having a birth in that spell.

@@

@@

Son preference in birth intervals shows most clearly for the fourth spell.
Despite the increase in spacing even for those with three girls, the increases in 
spacing have been substantially stronger for families with at least one son.
As for fertility, the more sons a family has, the more significant are the differences.

@@

Figure \ref{fig:bootstrap_med} shows that women with one to seven years of education 
follow a fertility pattern broadly similar to those with no education.
As expected, fertility is lower than for women with no education, but there is still a
strong son preference in the parity progression behavior.
Except for the second spell, the likelihood of having a birth by the end of the spell is
substantially lower if there has been at least one male birth than if there are only girls.
Despite this, the probability of a birth in the third and fourth spells has also been 
falling for those with only girls.
Having a second birth is still close to universal, although in urban areas it is down to
just over 85 percent for those with a son as their first child.

@@

The rapid decline in fertility for the most educated women is even more clear for the 
third and fourth spells when the woman has given birth to at least one son.
At the beginning of the sample, urban women with at least one boy had a close to 75
percent likelihood of having a birth, but that has fallen to only around 25 percent.
Son preference still shows clearly in the parity progression probabilities for 
women without a son, with the declines for the third and fourth spells substantially
slower for both urban and rural women without a son than for those with at least one son.
The falling fertility means that relatively few women make it to the fourth spell,
which, therefore, has more noise than the other spells.

@@

The tremendous impact that sex selection can have on birth spacing is illustrated 
particularly vividly by the changes over time for the third spell, which led to a complete 
reversal of the standard spacing pattern.
In the 1972-1984 period, the predicted average birth intervals for the three possible sex 
compositions were less than two months apart.
For the later periods, however, the highest predicted average spacing was for women
\emph{with only daughters}, with most of the differences to the other sex compositions 
statistically significant.
In urban areas, the predicted average birth interval for a woman with only daughters 
increased by almost a full year from 26.8 to 38.3 months.
What is more, this increase is likely an underestimate.
The sex ratios for the 1972-84 period are substantially higher than the natural sex ratio 
despite sex selection not being widely available until around 1985.
If this early, elevated sex ratio is because of recall error, it means that the actual birth 
intervals would have been substantially shorter than the ones predicted here and the
increase over time larger.

The reversal pattern also shows for the fourth spell for urban women, although not for
rural women.
An important caveat is that because of the falling fertility relatively few women make
it to the fourth spell, and even if they do, we often do not observe a birth by the time 
of the survey.
Of the 11,886 women across urban and rural areas observed for the fourth spell for the 
2005-2016 period more than 70 percent did not have a birth at the time of the survey.
Even though the censoring is less prevalent for the earlier periods, this is countered by
those samples being substantially smaller.
Despite this, the pattern for the fourth spell is similar to that of the third spell.

@@

The results above show consistent increases in average spacing across the board.
Average spacing is a convenient way to understand the overall changes in behavior but may 
hide important differences in the distribution of spacing.
Specifically, whether these increases in average spacing come from fewer births after 
very short intervals, because of even longer spacing from births with already long 
spacing, or a general increase in spacing have important implications for what effects the
increases in average spacing are likely to have and how to interpret the motivation for
the changes.
On one hand, if there still are many births that occur after very short intervals, the 
increase in average spacing is unlikely to have a substantial impact on health outcomes
since the literature on birth spacing and child health suggests that an interpregnancy 
interval of fewer than 18 months is associated with a higher risk of adverse health 
outcomes and mortality \citep{Conde-Agudelo2012}.
On the other hand, we may see a reduction in early observed births and more long spells if 
families with fewer sons are both more likely to have a very early pregnancy and to 
continue through pregnancies and sex-selective abortions until they conceive a son.%
\footnote{
It is unclear what the optimal response in the timing of pregnancies to the availability 
of sex selection should be.
The expected birth intervals become longer with the use of sex selection, which may be 
costly for parents leading them to try to conceive even earlier than previously.
Conversely, if parents know that short birth intervals are detrimental to the next child's
health, and that, with sex selection, the next child is likely a boy, it might make sense to 
wait longer before conceiving.
}

@@

Figure \ref{fig:pps_low} shows the distribution of birth within a spell for women with no 
education.
For the second spell, what is most interesting is that there is almost no difference 
between those with a first-born boy and those with a first-born girl.
For the 1972--1984 period the two conditional survival curves only begin to diverge 
at around 18 months (corresponding to 27 months after the first birth took place), and
at that point close to 50 percent of the women have already had their second child.
With a nine-month pregnancy, this means that almost 50 percent of the children born are
born within the interpregnancy interval associated with a higher risk of adverse health
outcomes and mortality, and, somewhat surprisingly, that there do not appear to be any 
particular disadvantage for the child born after a girl than after a boy.%
\footnote{
Note that this does not imply that the first-born girl is not in a more adverse position
than a first-born boy when the second child arrives, just that this is not directly
related to a short birth interval.
}
The difference in spacing after a first-born girl and a first-born boy remains almost 
the same across the four decades.
Furthermore, the proportion with very short spacing remains almost the same.

For the third spell, the son preference driven short spacing is evident throughout the
distribution and becomes more pronounced over the four decades.
Families with two girls are especially likely to have a birth early with 25 percent
having a child within 12 months of the beginning of the spell (21 months after the second 
birth).
That number is only slightly smaller four decades later. 
The fourth spell shows changes in the distribution that appear to be more pronounced
than for the third spell, although these results are based on fewer observed births.
For families with three girls, 75 percent have had their fourth child within 36 months 
after the beginning of the spell, but this occurs more than a year later for families
with two or more boys.

@@

For the third spell, the 1972--1984 period shows a close to standard pattern throughout, 
although, as mentioned above, the sex ratios for women without girls was substantially 
higher than the natural sex ratio either because of recall error or early use of 
sex selection.
The correlation between the increased use of sex selection and the reversal of the 
standard spacing pattern show up clearly in the 2005--2016 period for the third spell with 
spacing to the third child consistently longer with two girls than with either two boys 
or one boy and one girl throughout the entire spell.
This pattern is most consistent with early and continued use of sex selection, rather 
than a general reduction in the likelihood of very short birth intervals.
It also means that the previously born girls can expect more time without a younger sibling,
which may reduce the sibling competition for resources.

@@

Figure \ref{fig:bootstrap_high_r2} shows the same outcomes for the ``North'' region. 
Despite the idea that the ``North'' has high fertility and lower 
likelihood of sex selection the differences to the ``West'' region are surprisingly small,
at least in the urban areas.
Fertility is, indeed, higher than in the ``West'' region, but the declines in the 
probability of parity progression have been relatively rapid and mirror the national one.
The use of sex selection is also close to the ``West,'' and there is the same reversal of
spacing pattern as there.
The rural areas of the ``North'' are a slightly different story.
Fertility is substantially higher, and the use of sex selection much lower than in the
rural areas of the ``West.'' 
That said, there still is clear evidence that the sex ratio is substantially above normal
at close to 60 percent for the latest decade.

@@

The final region is the ``South,'' shown in Figure \ref{fig:bootstrap_high_r4}.
This region is historically thought to have the lowest degree of son preference of the
regions of India.
As for the ``East,'' I do not show the fourth spell results because of the small sample 
size and the associated noisy estimates.
Judging from the second spell progression rate and the associated spacing pattern, it
does, initially, appear that there is little son preference in this region.
The likelihood of having a second birth is relatively high at around 80 to 90 percent
and is close to independent of the sex of the first birth in both urban and rural areas.
The predicted average spacing also show no sign of son preference, and the predicted
sex ratios are close to the natural rate.
The average birth spacing started high relative to the other regions, and with little 
change in fertility and no significant use of sex selection, it has increased only 
slightly over time.

Despite the second spell's apparent lack of son preference behavior, for the third spell 
in urban areas the average birth interval has increased substantially for those with only 
girls relative to those with a boy and a girl, corresponding to
a predicted sex ratio close to 60 percent with only girls. 
The signs of the use of sex selection also show up for rural women with no sons.
Furthermore, over the four decades, the likelihood of having a third birth has decreased
much less than for either a boy and a girl or two boys.%
\footnote{
The relatively high parity progression for women without son does raise the question of
whether ``daughter avoidance'' is behind the fertility behavior in this region as
suggested by \citet{DiamondSmith2008}.
}
The differences in parity progression and the sex ratios suggest that despite originally
having less son preference, the ``South'' now appears to have longer average spacing 
in the absence of sons and the higher sex ratio associated with the use of sex selection.

@@

At the other extreme, urban women with eight or more years of education have experienced 
an almost complete reversal of the traditional spacing patterns as a result of falling
fertility and the associated extensive use of sex selection.
Women with either no or one son now have substantially longer spacing than if they have 
two or more sons, rather than having the next birth sooner as they mostly did before sex 
selection became available.
As shown by the increasingly significant deviations in the sex ratio from the natural sex 
ratio, these changes are due to increased use of sex selection rather than a sudden
reversal in son preference.
Only the second spell has not seen a reversal despite a substantially increased use of
sex selection, mainly because the average spacing with a first-born boy has increased 
independently by more than nine months.
